---
title: Integrated evaluation of clinical, pathological and radiological prognostic factors in squamous cell carcinoma of the lung
author: Lisa KOPPE, Khwansiri NINPAN, Princy PAPPACHAN
date: 4/23/2020
---

Import the necessary packages:

```{r}
library(tidyverse)
library(survival)
library(ggplot2)
```

Load the lung cancer dataset:

```{r}
raw_data <- read.csv("lungcancer.csv", header=TRUE, sep=";", 
                     col_types=cols(Index='',
                                    Sex='',
                                    Age='',
                                    Location='',
                                    Obst_pn_or_plugging='',
                                    CT_value_Mean='',
                                    CT_value_Std_Dev='',
                                    CT_value_P_major='',
                                    CT_V_ratio='',
                                    CT_Kernel='',
                                    Necrosis='',
                                    Underlying_lung='',
                                    Multiplicity='',
                                    Effusion='',
                                    Bronchoscopy='',
                                    Differentiation='',
                                    pos_LN='',
                                    total_LN='',
                                    LN_ratio='',
                                    cT='',
                                    cN='',
                                    cM='',
                                    pT='',
                                    pN='',
                                    pM='',
                                    Smoking_state='',
                                    PY='',
                                    Cessation_time='',
                                    FVCpercentPRED='',
                                    FEV1percentPRED='',
                                    FEV1_FVC='',
                                    CEA='',
                                    post_opCTx='',
                                    post_op_RTx='',
                                    op_type='',
                                    death='',
                                    recur_or_meta='',
                                    OP_site_recur='',
                                    meta='',
                                    meta_site='',
                                    OS='',
                                    DFS=''))
```

```{r}
data <- as_tibble(raw_data)
summary(data)
names(data)
str(data)
dim(data)
head(data)
```

```{r}

```

# Univariate analysis

Global columns:
- Location: peripheral vs. central (cat)
- death: die vs. survive (bin)
- OS: overall survival: interval between the date of diagnosis and date of death (by any cause)(cont)

-----

Binary vs Binary: Fisher Test
Binary vs Continuous: Logistic Regression

## Univariate analysis - Lisa:

Clinicopathologic (part2):
- FEV1_FVC: pulmonary function test: forced expiratory volume in a second / forced vital capacity (cont)
- CEA: serum carcinoembryonic antigen (tumor marker) (cont)
- LN_ratio: lymph node ratio: number of nodes with positive tumor cells divided by the number of all resected nodes (cont)
- meta: metastasis observed (bin)
- pT: peripheral / size of original tumor (cont)
- pN: peripheral / nearby lymph nodes involved (cont)
- pM: peripheral / distant metastasis (spread of cancer from one part of the body to another) (cont)

### Binary outcome:


```{r}

```


