---
title: "Lung Cancer Prognosis"
author: "Lisa KOPPE, Khwansiri NINPAN, Princy PAPPACHAN"
date: "4/23/2020"
output: word_document
---

Data Source:
[Integrated evaluation of clinical, pathological and radiological prognostic factors in squamous cell carcinoma of the lung](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6777828/)

# Import the necessary packages:

```{r}
library(tidyverse)
library(survival)
library(survminer)
```

# Data preparation: load the lung cancer dataset and check data types

```{r}

# Check data types only columns we want to work with
# Set default as c for the rest

lungSCC_raw <- read_csv("LungCancer.csv",
                         col_types=cols(Sex='c',
                                       Age='d',
                                       Location='d',
                                       Obst_pn_or_plugging='d',
                                       Necrosis='d',
                                       Underlying_lung='d',
                                       Effusion='d',
                                       LN_ratio='d',
                                       cT='d',
                                       cN='d',
                                       cM='d',
                                       pT='d',
                                       pN='d',
                                       pM='d',
                                       Smoking_state='d',  
                                       FEV1_FVC='d',
                                       CEA='d',
                                       op_type='d',
                                       death='d',
                                       meta='d',
                                       OS = "d",
                                       .default = 'c'))
```

# Explore the dataset:
    - For classification data, use "table"
    - For numerical data: use "table" + "summary"
      (Summary to define basic descriptive statistic, Table to check null value)

```{r}
dim(lungSCC_raw)  # Check total patients + columns


# Radiologic features
table(lungSCC_raw$Location, useNA="always")
table(lungSCC_raw$Obst_pn_or_plugging, useNA="always")
table(lungSCC_raw$Necrosis, useNA="always")
table(lungSCC_raw$Underlying_lung, useNA="always")
table(lungSCC_raw$Effusion, useNA="always")

# Clinicopathologic features
table(lungSCC_raw$Sex, useNA="always")
table(lungSCC_raw$Age, useNA="always")
summary(lungSCC_raw$Age)
table(lungSCC_raw$LN_ratio, useNA="always")
summary(lungSCC_raw$LN_ratio)
table(lungSCC_raw$cT, useNA="always")
table(lungSCC_raw$cN, useNA="always")
table(lungSCC_raw$cM, useNA="always")
table(lungSCC_raw$pT, useNA="always")
table(lungSCC_raw$pN, useNA="always")
table(lungSCC_raw$pM, useNA="always")
table(lungSCC_raw$Smoking_state, useNA="always")
table(lungSCC_raw$FEV1_FVC, useNA="always")
summary(lungSCC_raw$FEV1_FVC)
table(lungSCC_raw$CEA)
summary(lungSCC_raw$CEA)
table(lungSCC_raw$op_type, useNA="always")
table(lungSCC_raw$death, useNA="always")
table(lungSCC_raw$meta, useNA="always")
table(lungSCC_raw$OS, useNA="always")
summary(lungSCC_raw$OS)
```


Redefine the meaning of each label to make it easier for interpretation

```{r}
lungSCC <- mutate(lungSCC_raw,
                  Location=ifelse(Location==0, "Peripheral","Central"),
                  
                  Necrosis = factor(Necrosis, levels=c("0","1","2"), labels = c("NoRemark", "Necrosis","Cavitation")),
                  Necrosis <- relevel(Necrosis, "NoRemark"),
              
                  Underlying_lung= factor(Underlying_lung, levels = c("0","1","2"), labels = c("NoRemark", "Emphysema", "ILA")),
                  Underlying_lung <- relevel(Underlying_lung, "NoRemark"),
                  
                  Sex=factor(Sex, levels=c("F", "M")),
                  
                  AG=ifelse((Age<60), "LT60", "OV60"),
                  AG=factor(AG),
                
                  Smoking_state=factor(Smoking_state, levels = c("3","0","1","2"), labels=c("Non-smoker", "Current-smoker", "Ex-smoker","Unknown" )),
                  Smoking_state <- relevel(Smoking_state, "Non-smoker"),
                  
                  OP = factor(op_type, levels = c("5","1", "2", "3", "4"), labels = c("No op", "Lobectomy", "Pneumonectomy", "Segmentectomy", "LN biopsy")),
                  OP <- relevel(OP, "No op"),
                  
                  cT=factor(cT, levels=c("1", "2", "3", "4"), labels=c("cT1", "cT2", "cT3", "cT4")),
                  
                  cN=factor(cN, levels=c("1", "2", "3", "4"), labels=c("cN0", "cN1", "cN2", "cN3")),
                  
                  cM=factor(cM, levels=c("1", "3"), labels=c("cM0", "cM1b")),
                  
                  pT=factor(pT, levels=c("1", "2", "3", "4"), labels=c("pT1", "pT2", "pT3", "pT4")),
                  
                  pN=factor(pN, levels=c("1", "2", "3", "4"), labels=c("pN0", "pN1", "pN2", "pN3")),
                  
                  pM=factor(pM, levels=c("1", "3"), labels=c("pM0", "pM1b")))

```


Recheck the distribution of age after regrouped
```{r}
table(lungSCC$AG)
```



# Features description

1. Location
   = Tumor location
   (0 = Peripheral SCC, 1 = Central SCC)

2. death
   (0 = Alive, 1 = Dead)

3. OS
   = Overall Survival in months: interval between the date of diagnosis and date of death (by any cause)

## Radiologic features:

4. Obst_pn_or_plugging
   = Presence of obstructive pneumonitis/atelectais
   (0 = Absent, 1 = Present)

5. Necrosis
   (0 = No Remark, 1 = Necrosis, 2 = Cavitization)

6. Underlying_lung
   = Underlying lung disease
   (0 = No Remark, 1 = Emphysema, 2 = Interstitial lung abnormality)

7. Effusion
   (0 = Absent, 1 = Present)

## Clinicopathologic features:

8. Sex
   = Gender of the patient
   (F=Female, M=Male)

9. AG
   = Age at diagnosis
   (LT60 = Less than 60 years old, OV60 = Over 60 years old)

10. LN_ratio
    = Lymph node ratio: Number of nodes with positive tumor cells divided by the number of all resected nodes

11. cT
    = Size of original tumor in central lung SCC patients
    (1 = T1, 2 = T2, 3 = T3, 4 = T4)

12. cN
    = Nearby lymph nodes involved in central lung SCC patients
    (1 = N0, 2 = N1, 3 = N2, 4 = N3)

13. cM
    = Distant metastasis (spread of cancer from one part of the body to another) in central lung SCC patients
    (1 = M0, 2 = M1a, 3 = M1b, 4 = M1ab, 5 = Mx)

14. pT
    = Size of original tumor in peripheral lung SCC patients
    (1 = T1, 2 = T2, 3 = T3, 4 = T4)

15. pN
    = Nearby lymph nodes involved in peripheral lung SCC patients
    (1 = N0, 2 = N1, 3 = N2, 4 = N3)

16. pM
    = Distant metastasis (spread of cancer from one part of the body to another) in peripheral lung SCC patients
    (1 = M0, 2 = M1a, 3 = M1b, 4 = M1ab, 5 = Mx)

17. Smoking_state
    = Current smoking state of the patient
    (current-smoker, ex-smoker, non-smoker, unknown)

18. FEV1_FVC
    = Pulmonary function test in %: forced expiratory volume in a second / forced vital capacity

19. CEA
    = Serum carcinoembryonic antigen level in ng/ml (tumor marker)

20. OP
    = Type of operation the person has undergone
    (lobectomy, pneumonectomy, segmentectomy, LN biopsy, no Op)

21. meta
    = Metastasis observed
    (0=Absent, 1=Present)


*Part 1: Test relationship between radiologic/clinicopathologic features and death*

Methods used:
1. Fisher's exact test (determine the relationship between two categorical variables)
2. Logistic regression (determine the relationship between categorical continuous variables)

## Relationship between radiologic features and death

### Fisher's exact test

Determine the relationship between two categorical variables

Hypotheses:
H0: there is no relationship between the two categorical variables.
H1: the variables are dependent. Knowing the value of one variable helps to predict the value of the other variable.


----
1. Difference in risk of death in patient with and without obstructive pneumonitis/ atelectais
2. Difference in risk of death in patient with and without lung effusion
----


1. Difference in risk of death in patient with and without obstructive pneumonitis/ atelectais:
```{r}
with(lungSCC, fisher.test(table(death, Obst_pn_or_plugging)))
```



2. Difference in risk of death in patient with and without lung effusion:
```{r}
with(lungSCC, fisher.test(table(death, Effusion)))
```


### Logistic regression

Determine the relationship between categorical continuous variables

Hypotheses:
H0: there is no different in odds between that specific group and reference group
H1: Odds of that specific group and reference group is statistically significant

Interpretation:
Estimate of interception = The log odds when covariate(s) equal 0
Estimate of each variable = The increase/decrese in the log odds when compare to reference group
To define the odds ratio, exponent estimate value.

----
1. Difference in risk of death based on cancer location
2. Difference in risk of death based on cell necrosis
3. Difference in risk of death based on underlying lung disease
----


1. Difference in risk of death based on cancer location:
```{r}
fit <- glm(death~Location, family="binomial", data=lungSCC)
summary(fit)
```


```{r}
exp(0.2749)
```


```{r}
exp(confint(glm(death ~ Location, data = lungSCC, family = "binomial"))[2,])
```



2. Difference in risk of death based on cell necrosis:
```{r}
fit <- glm(death~Necrosis, family="binomial", data=lungSCC)
summary(fit)
```



```{r}
exp(-0.7482)
exp(-0.9776)
```



3. Difference in risk of death based on underlying lung disease:
```{r}
fit <- glm(death~Underlying_lung, family="binomial", data=lungSCC)
summary(fit)
```


```{r}
exp(1.76823)
exp(-0.51250)
```



## Relationship between clinicopathologic features and death

### Fisher's exact test

Determine the relationship between two categorical variables

Hypotheses:
H0: there is no relationship between the two categorical variables.
H1: the variables are dependent. Knowing the value of one variable helps to predict the value of the other variable.

Odds ratio:
Odds for death in patient with specific condition
divided by odds for death in patient without specific condition
If those 2 sets of patient have equal odds (odds ratio = 1), that specific condition are independent to death status

----
1. Difference in risk of death in patient based on gender
2. Difference in risk of death in patient with an age lower or greater than 60
3. Difference in risk of death in patient with and without observed metastasis after surgery (meta)
----


1. Difference in risk of death in patient based on gender:
```{r}
with(lungSCC, fisher.test(table(death, Sex)))
```



2. Difference in risk of death in patient with Age less than and over 60:
```{r}
with(lungSCC, fisher.test(table(death, AG)))
```



3. Difference in risk of death in patient with and without observed metastasis after surgery (meta):
```{r}
with(lungSCC, table(death, meta))
with(lungSCC, fisher.test(table(death, meta)))
```


### Logistic regression

Determine the relationship between categorical continuous variables

----
1. Difference in risk of death based on lymph node ratio
2. Difference in risk of death based on cT Central Lung SCC size of Tumor
3. Difference in risk of death based on cN when nearby lymph nodes are involved
4. Difference in risk of death based on cM when distant metastasis are involved
5. Difference in risk of death based on pT Peripheral Lung SCC size of Tumor
6. Difference in risk of death based on pN when nearby lymph nodes are involved
7. Difference in risk of death based on pM when distant metastasis are involved
8. Difference in risk of death based on the smoking state
9. Difference in risk of death based on FEV1_FVC pulmonary function test
10. Difference in risk of death based on CEA serum
11. Difference in risk of death based on the type of operation undergone
----

1. Difference in risk of death based on lymph node ratio:
```{r}
with(lungSCC, table(death, LN_ratio))

fit <- glm(death~LN_ratio, family="binomial", data=lungSCC)
summary(fit)

exp(confint(glm(death~LN_ratio, family="binomial", data=lungSCC))[2,])
```


```{r}
exp(1.38)
```



2. Difference in risk of death based on cT Central Lung SCC size of Tumor:
```{r}
summary(glm(death ~ cT, data = lungSCC, family = "binomial"))
```

```{r}
exp(0.5702)
exp(1.4731)
exp(1.3122)
```



3. Difference in risk of death based on cN when nearby lymph nodes are involved:
```{r}
summary(glm(death ~ cN, data = lungSCC, family = "binomial"))
```

```{r}
exp(0.5702)
exp(1.4731)
exp(1.3122)
```


4. Difference in risk of death based on cM when distant metastasis (spread of cancer from one part of the body to another) are involved:
```{r}
summary(glm(death ~ cM, data = lungSCC, family = "binomial"))
```

```{r}
exp(0.2744)
```



5. Difference in risk of death based on pT Peripheral Lung SCC size of Tumor:
```{r}
with(lungSCC, table(death, pT))

fit <- glm(death~pT, family="binomial", data=lungSCC)
summary(fit)

exp(confint(glm(death~pT, family="binomial", data=lungSCC))[2,])
```

```{r}
exp(0.4370)
exp(0.9622)
exp(1.2672)
```


6. Difference in risk of death based on pN when nearby lymph nodes are involved:
```{r}
with(lungSCC, table(death, pN))

fit <- glm(death~pN, family="binomial", data=lungSCC)
summary(fit)

exp(confint(glm(death~pN, family="binomial", data=lungSCC))[2,])
```

```{r}
exp(0.2224)
exp(1.1628)
exp(1.5965)
```


4. Difference in risk of death based on pM when distant metastasis (spread of cancer from one part of the body to another) are involved:
```{r}
with(lungSCC, table(death, pM))

fit <- glm(death~pM, family="binomial", data=lungSCC)
summary(fit)

exp(confint(glm(death~pM, family="binomial", data=lungSCC))[2,])
```

```{r}
exp(15.8626)
```


8. Difference in risk of death based on the smoking state:
```{r}
summary(glm(death ~ Smoking_state, data = lungSCC, family = "binomial"))
```


9. Difference in risk of death based on FEV1_FVC pulmonary function test:
```{r}
with(lungSCC, table(death, FEV1_FVC))

fit <- glm(death~FEV1_FVC, family="binomial", data=lungSCC)
summary(fit)

exp(confint(glm(death~FEV1_FVC, family="binomial", data=lungSCC))[2,])
```

```{r}
exp(0.011147)
```


10. Difference in risk of death based on CEA serum:
```{r}
with(lungSCC, table(death, CEA))

fit <- glm(death~CEA, family="binomial", data=lungSCC)
summary(fit)

exp(confint(glm(death~CEA, family="binomial", data=lungSCC))[2,])
```

```{r}
exp(-0.003081)
```


11. Difference in risk of death based on the type of operation undergone:
```{r}
summary(glm(death ~ OP, data = lungSCC, family = "binomial"))

```

```{r}
exp(2.0974)

```



