---
title: "LungCancer_Multivar"
author: "Khwansiri NINPAN"
date: "27/04/2020"
output:
  word_document: default
  html_document:
    df_print: paged
---

*Part 3: Part 3: Model building and diagnosis*

## Multivaraites Cox Regression

```{r}
library(survival)
library(tidyverse)
library(survminer)
library(asaur)

```

```{r}
lungSCC_raw <- read_csv("LungCancer.csv",
                         col_types=cols(Sex='c',
                                       Age='c',
                                       Location='d',
                                       Obst_pn_or_plugging='d',
                                       Necrosis='d',
                                       Underlying_lung='d',
                                       Effusion='d',
                                       LN_ratio='d',
                                       cT='c',
                                       cN='c',
                                       cM='c',
                                       pT='c',
                                       pN='c',
                                       pM='c',
                                       Smoking_state='c',  
                                       FEV1_FVC='d',
                                       CEA='d',
                                       op_type='c',
                                       death='d',
                                       OP_site_recur='l',
                                       meta='l',
                                       OS = "d",
                                       .default = 'c'))


lungSCC <- mutate(lungSCC_raw,
                  Location=ifelse(Location==0, "Peripheral","Central"),
                  
                  Necrosis=ifelse(Necrosis==0, "NoRemark",
                                  ifelse(Necrosis==1, "Necrosis", "Cavitization")),
                  
                  Underlying_lung=ifelse(Underlying_lung==0, "NoRemark",
                                         ifelse(Underlying_lung==1, "Emphysema", "ILA")),
                  
                  Sex=factor(Sex, levels=c("F", "M")),
                  
                  AG=ifelse((Age<60), "LT60", "OV60"),
          
                  AG=factor(AG),
                
                  OP=factor(op_type, labels=c("lobectomy", "pneumonectomy", "segmentectomy", "LN biopsy", "No OP")),
                                                  
                  Smoking_state=factor(Smoking_state, labels=c("unknown", "current-smoker", "ex- smoker", "non- smoker")),
                  
                  cT=factor(cT, levels=c("1", "2", "3", "4"), labels=c("cT1", "cT2", "cT3", "cT4")),
                  
                  cN=factor(cN, levels=c("1", "2", "3", "4"), labels=c("cN0", "cN1", "cN2", "cN3")),
                  
                  cM=factor(cM, levels=c("1", "3"), labels=c("cM0", "cM1b")),
                  
                  pT=factor(pT, levels=c("1", "2", "3", "4"), labels=c("pT1", "pT2", "pT3", "pT4")),
                  
                  pN=factor(pN, levels=c("1", "2", "3", "4"), labels=c("pN0", "pN1", "pN2", "pN3")),
                  
                  pM=factor(pM, levels=c("1", "3"), labels=c("pM0", "pM1b")))


```



Start with the full model with all variables
```{r}

lungSCC_fullmodel <- coxph(Surv(OS, death) ~ Location + Obst_pn_or_plugging + Necrosis + Underlying_lung + Effusion + Sex + AG + LN_ratio + cT+ cN + cM + pT+ pN+ pM + Smoking_state + FEV1_FVC + CEA + meta , data = lungSCC)

```


Automatic model selection based on AIC
``` {r}
lungSCC_multiCox <- step(lungSCC_fullmodel)
```

Note that this backward selection migh not convert to global optimum

```{r}
summary(lungSCC_multiCox)
confint(lungSCC_multiCox)  # 95% CI
```



#Predictive power: AUC
Delete death and OS information and make the prediction by using our model
```{r}
library(pROC)
lungSCC_predict <- select(lungSCC, -death, -OS)
lungSCC_predict$lp <- predict(lungSCC_multiCox, newdata=lungSCC_predict, type="lp")


roc(lungSCC$death,lungSCC_predict$lp, 
    smoothed = TRUE,
    ci = TRUE, ci.alpha = 0.9, stratified = FALSE,
    plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, grid = TRUE,
    print.auc=TRUE, show.thres=TRUE)

```


Model diagnostics: Case deletion residuals
```{r}
dfbetas <- residuals(lungSCC_multiCox, type = 'dfbetas')
lungSCC$dfbetas <- sqrt(rowSums(dfbetas^2))
plot(lungSCC$dfbetas, type = 'h')
abline(h = 0)
```
